### 2020-07-03 manuscript/manuscript-speed-accuracy.tex
Updated README. Re-running analysis.

We found that author reputation, journal impact, the number of citations, software speed and age are not reliable predictors of software accuracy. This implies that useful bioinformatics software is not only the domain of famous researchers, and that any researchers are capable of producing good software. In addition, we found that there exists an excessive number of slow and inaccurate software tools across multiple sub-disciplines of bioinformatics. Meanwhile, there are very few tools of middling accuracy and speed. We hypothesise that a strong publication bias is unduly influencing the publication and development of bioinformatic software tools. In other words, at present software that is not highly ranked on speed and not highly ranked on accuracy is difficult to publish due to editorial and reviewer practices. This leaves an unfortunate gap in the literature upon which future software refinements could be constructed.
}

@@

\section*{Background}
Computational biology software is widely used and has produced some of the most cited publications in the scientific corpus \cite{Perez-Iratxeta2007-lv,Van_Noorden2014-kc,Wren2016-xy}. This software includes implementations of methods for sequence alignment and homology inference \cite{Altschul1990-ht,Thompson1994-eu,Thompson1997-rl,Altschul1997-ga}, phylogenetic analysis \cite{Felsenstein1985-lj,Saitou1987-zl,Posada1998-qq,Ronquist2003-yh,Tamura2007-ei}, statistical analysis of survival patterns in biomedicine \cite{Kaplan1958-ju,Cox1972-nu}, biomolecular structure analysis \cite{Sheldrick1990-kc,Sheldrick2008-xy,Jones1991-ik,Laskowski1993-vi,Otwinowski1997-xj}, visualization and data collection \cite{Kraulis1991-lt,Berman2000-to}. However, the popularity of a software tool does not necessarily imply that it is either accurate or computationally efficient, instead usability, ease of installation, operating system and other factors may play a greater role.

@@

Other factors that influence whether a software tool is selected include: whether the documentation is good, user-friendly, word-of-mouth and ``used in a similar analysis'' \cite{Loman2015-bw}. This sort of information is not as readily quantifiable as the above measures. However, citation metrics may be a useful proxy. The word-of-mouth factor may also explain the reason why some software continues to be used, in spite of poor relative performance \cite{Wadi2016-dj}.

@@

For each of the publications describing these methods, we have (when possible) identified the 2014 journal impact factor (JIF), published by Thomson Reuters \cite{Garfield2006-lv} and the H5-index published by \href{https://scholar.google.co.nz/citations?view_op=top_venues&hl=en}{Google Scholar Metrics}. We have collected the H-indices and M-indices \cite{Hirsch2005-mt} for the corresponding authors for each method, and the number of times the publication(s) associated with a method has been cited using Google Scholar (data collected over a 1 month period in early 2016).

@@

In order to gain a deeper understanding of the distribution of available bioinformatic software tools on a speed versus accuracy landscape, we ran a Monte Carlo permutation test. The ranks extracted from each benchmark were randomly permuted, generating 10,000 randomized speed and accuracy ranks. In the cells of a $10\times10$ grid spanning the normalised speed and accuracy ranks we computed a Z-score for the observed number of methods in a cell, compared to the expected distributions generated by the randomized ranks. The results of which are shown in Figure~\ref{fig:speedaccuracy}. We identified 21 bins where there was a significant excess or dearth of methods. For example, there was an excess of “slow and inaccurate” software (Z=1.6, P-value=0.05) and “slow and accurate” software (Z=1.9, P-value=0.03). We find that the amount of software classed as “fast and accurate” and “fast and inaccurate” are at approximately the expected proportions based upon the permutation test. The number of significant results is in excess and not due to multiple testing, as the probability of finding 21 of 100 tests significant by chance is low (P-value = $2\times 10^{-8}$, exact binomial test) \cite{Moran2003-ve}.

@@

We have gathered data on the relative speed and accuracies of \textbf{{\color{black}243}} bioinformatic methods from \textbf{{\color{black}43}} benchmarks that were published between 2005 and 2016. We show that there is an under-representation of software that has both intermediate levels of accuracy and speed. There may be a number of factors that drive this phenomena. One likely explanation is that bioinformatic software tools suffer from a form of publication bias \cite{Boulesteix2015-am}. Our community of developers, reviewers and editors may be unwilling to publish software that is neither the fastest nor the most accurate (we have anecdotal evidence to support this, N=1). If correct, this is unfortunate. Some problems that lack fast and accurate solutions force researchers to make unnecessary compromises in terms of accuracy and time. If our hypothesis that this underrepresentation is due to publication bias is valid then why is there an enrichment of slow and inaccurate software (P-value=0.05, empirical distributions from permutation tests)? How could these methods be published? We have found that slow and inaccurate software is generally published earlier than alternative methods (P=0.007, one-tailed Wilcoxon test), therefore comparisons were not required to publish these tools and the accuracies may have been over-optimistically reported \cite{Boulesteix2015-nc}.

We found that no commonly used proxy for study quality is correlated with software accuracy. Neither, author reputation, number of citations, journal impact, relative age or speed appear to be associated with accurate software. Linear mixed models of these values also fail to identify predictors of software accuracy, as do methods for combining P-values. A great deal of criticism has been leveled specifically at journal impact factors \cite{Editors2006-ue,Rossner2007-yk,Lariviere2010-kx}. Our finding show that this measure is not reflective of software quality. But neither is the H5 index which is thought to be a more robust measures of journal impact \cite{Harzing2008-wb}. The poor relationship between accuracy and both author reputation and the number of citations is particularly troubling. Both are related to “word of mouth” and “previously used in a similar analysis” which a recent survey of researchers suggested is a major influence on the selection of software tools \cite{Loman2015-bw}. This implies that the recorded high citation rates for bioinformatic software  \cite{Perez-Iratxeta2007-lv,Van_Noorden2014-kc,Wren2016-xy} is more a reflection of user-friendliness and the Matthew Effect \cite{Lariviere2010-kx,Merton1968-cb}. The Matthew Effect is a biblical reference that can be paraphrased as “the rich get richer and the poor get poorer”. In this context, highly cited software is more likely to be used and cited again, irrespective of relative performance.

The lack of any relationship between software speed and accuracy is surprising. 
The slower software tools are found to be overrepresented at both high and low levels of accuracy (Figure~\ref{fig:speedaccuracy}). Likewise accurate software was found in both high and low speed ranges. A simple gedankenexperiment may be sufficient to prove that slow software is less thoroughly tested than fast software. This is because typical software development is an iterative process, where methods are refined by successive rounds of coding, testing and evaluation \cite{Wilson2006-ih}. We can assume that similar time-spans are spent on most projects, for example, the span of a MSc or PhD degree.  As a consequence slow software undergoes fewer development cycles than fast methods, and is therefore tested less, often resulting in less accurate and slower software. The fact that fast and inaccurate software is relatively rare does support this argument. The lack of any further apparent relationship between speed and accuracy implies there are stronger influences on accuracy rankings than speed alone.

\section*{Conclusions}
Scientific progress is a grinding process made by testing hypotheses using appropriate positive and negative controls, in combination with independent experiments and replication \cite{Ioannidis2005-xh,Moonesinghe2007-kk}. We think that software and analysis tools should be no exception to these principles. The continual increases in the complexity of software tools creates an increased likelihood that software bugs are introduced \cite{Darriba2015-sz}. Scientific software should therefore be thoroughly tested by developers, however, the results of developer and author-derived tests should be treated with caution due to a range of potential conscious and unconscious biases \cite{Norel2011-cq}. Our study shows that no commonly-used impact-based metrics are related to software accuracy, therefore the only reasonable way to select software tools is through software benchmarks.

We have shown that slow and inaccurate software is typically published early in the development of a field (see Figure S8). Subsequent software tools may be high accuracy or high speed. However, as further software tools become available, those that do not rank highly in terms of accuracy or speed are under-represented in the literature. Our conjecture is that this is due to a publication bias in computational biology software literature, with publishers, editors and reviewers implicitly requiring software to be notable in at least one of these regards in order to be published. This hole in the literature leaves an unfortunate gap upon which further software refinements cannot be made.

The software that could be used in building research workflows can make use of software tools that are of medium accuracy and speed. These may prove to be useful rapid pre-filters for more accurate, yet computationally-demanding approaches. These could also be used when datasets are too large to be processed by more accurate software tools, i.e. when accurate tools are slow or memory hungry \cite{Lowe1997-hq,Weinberg2006-tm}. These tools may also serve as an indication that some approaches may not be worth pursuing further.

We propose that the full spectrum of software tool accuracies and speeds serve a useful purpose to the research community. Like negative results, if honestly reported, illustrate to the research community that certain approaches are not practical research avenues \cite{Ioannidis2005-xh,Workman1999-au,Rivas2000-fb}. The classes of tools that we find in the under-represented accuracy and speed region include some of the most used tools in bioinformatics, such as the homology search tools NCBI-BLAST \cite{Altschul1990-ht,Altschul1997-ga} and HMMER2 \cite{Eddy:2009} and genome assembly and mapping tools such as AbySS \cite{Simpson:2009}, MAQ \cite{Li:2008} and Newbler \cite{Margulies:2005}. The current practises of publishers, editors, reviewers and authors of software tools therefore deprive our community of tools for building effective and productive workflows.

@@

We have shown that accurate software is not necessarily the most recently released method or the product of high profile lab groups or selected by high impact journals. Software that is widely used or is either slow or fast is also not necessarily the most accurate. Therefore, accurate software may be the product of features we have not been able to capture. Possibly, hard work, good ideas and sound method testing, as well as technical ability, experience and education levels in software development \cite{Wilson2006-ih}.

Finally, we think that the field of computational biology could benefit from embracing an increased number of independent software comparison studies \cite{Boulesteix2013-vb}, in parallel with the popular challenge-based benchmarks such as CASP, Assemblathon, Alignathon, DREAM and RGASP \cite{Boutros2014-zm}. We are hopeful that this, along with the relaxation of the publication bias we have described, will reduce the over-optimistic and misleading reporting of method accuracy \cite{Boulesteix2010-te,Jelizarow2010-zf,Norel2011-cq}.

@@

\section*{Acknowledgements}
The authors acknowledge the valued contribution of invaluable discussions with Anne-Laure Boulesteix, Shinichi Nakagawa, Suetonia Palmer and Jason Tylianakis. Michael A. Black,  Murray Cox, Raquel Norel, Alexandros Stamatakis, Jens Stoye,Tandy Warnow, provided valuable feedback on drafts of the manuscript.

PPG, FAG and NEW are supported by a Rutherford Discovery Fellowship,
administered by the Royal Society of New Zealand. NEW is supported by
a PhD scholarship from the University of Canterbury.

>>>>

We found that author reputation, journal impact, the number of citations, software speed and age are not reliable predictors of software accuracy. Unfortunately, citations, author and journal reputation are frequently cited reasons for selecting software tools. Our results imply that useful bioinformatics software is not only the domain of famous senior researchers. In addition, we found that there is an excess of slow and inaccurate software tools across multiple sub-disciplines of bioinformatics. Meanwhile, there are very few tools of middling accuracy and speed. We hypothesise that a form of publication-bias unduly influences the publication and development of bioinformatic software tools. In other words, at present software that is not highly ranked on speed and not highly ranked on accuracy is difficult to publish due to editorial and reviewer practices. This leaves an unfortunate gap in the literature upon which future software refinements cannot be constructed.
}

@@

%%REFERENCES::
%reference Stefan Buchka$^{1}$, Rory Wilson$^2$, Alexander Hapfelmeier$^3$,\\  Paul Gardner$^4$,  Anne-Laure Boulesteix PAPER!!!

%% A comprehensive analysis of the usability and archival stability of omics computational tools and resources
%% https://www.biorxiv.org/content/10.1101/452532v1

%% Tools and best practices for data processing in allelic expression analysis
%% https://genomebiology.biomedcentral.com/articles/10.1186/s13059-015-0762-6

%% Towards reproducible, transparent, and systematic benchmarking of omics computational tools
%% https://osf.io/p8yd9

%% Essential guidelines for computational method benchmarking
%% https://arxiv.org/abs/1812.00661

\section*{Background}
Computational biology software is widely used and has produced some of
the most cited publications in the scientific corpus
\cite{Perez-Iratxeta2007-lv,Van_Noorden2014-kc,Wren2016-xy}. This
software includes implementations of methods for sequence alignment
and homology inference
\cite{Altschul1990-ht,Thompson1994-eu,Thompson1997-rl,Altschul1997-ga},
phylogenetic analysis
\cite{Felsenstein1985-lj,Saitou1987-zl,Posada1998-qq,Ronquist2003-yh,Tamura2007-ei},
statistical analysis of survival patterns in biomedicine
\cite{Kaplan1958-ju,Cox1972-nu}, biomolecular structure analysis
\cite{Sheldrick1990-kc,Sheldrick2008-xy,Jones1991-ik,Laskowski1993-vi,Otwinowski1997-xj},
visualization and data collection
\cite{Kraulis1991-lt,Berman2000-to}. However, the popularity of a
software tool does not necessarily imply that it is either accurate or
computationally efficient, instead usability, ease of installation,
operating system and other factors may play a greater role. There have
been several notable incidences where inaccurate software has caused
considerable harm \cite{leveson1993investigation}.
%Fergusson covid predictions?

@@

Other factors that are less quantifiable that influence whether a
software tool is selected include: whether the documentation is good,
user-friendly, word-of-mouth and ``used in a similar analysis''
\cite{Loman2015-bw}. This sort of information is not as readily
quantifiable as the above measures. However, citation metrics may be a
useful proxy. The word-of-mouth factor may also explain the reason why
some software continues to be used, in spite of poor relative
performance \cite{Wadi2016-dj}. We note that none of these factors
are directly linked to the top ranked features of successful software
projects, as determined by literature survey \cite{nasir2011critical}.

{\bf EDITS:\\
  Number of releases. GitHub statistics (https://www.frontiersin.org/articles/10.3389/fbioe.2018.00198/abstract).\\
  The influence of organizational structure on software quality https://ieeexplore.ieee.org/iel5/4814109/4814110/04814163.pdf
  }

@@

For each of the publications describing these methods, we have (when possible) identified the 2014 journal impact factor (JIF), published by Thomson Reuters \cite{Garfield2006-lv} and the H5-index published by \href{https://scholar.google.co.nz/citations?view_op=top_venues\&hl=en}{Google Scholar Metrics}. We have collected the H-indices and M-indices \cite{Hirsch2005-mt} for the corresponding authors for each method, and the number of times the publication(s) associated with a method has been cited using Google Scholar (data collected over a 1 month period in early 2016).

@@

In order to gain a deeper understanding of the distribution of available bioinformatic software tools on a speed versus accuracy landscape, we ran a Monte Carlo permutation test. The ranks extracted from each benchmark were randomly permuted, generating 10,000 randomized speed and accuracy ranks. In the cells of a $10\times10$ grid spanning the normalised speed and accuracy ranks we computed a Z-score for the observed number of methods in a cell, compared to the expected distributions generated by the randomized ranks. The results of which are shown in Figure~\ref{fig:speedaccuracy}. We identified 19 bins where there was a significant excess or dearth of methods. For example, there was an excess of “slow and inaccurate” software (Z=1.6, P-value=0.05) and “slow and accurate” software (Z=1.9, P-value=0.03). We find that the amount of software classed as “fast and accurate” and “fast and inaccurate” are at approximately the expected proportions based upon the permutation test. The number of significant results is in excess and not due to multiple testing, as the probability of finding 19 of 100 tests significant by chance is low (P-value = $5\times 10^{-7}$, exact binomial test).

% binom.test(sigCount,gridRes^2, p = 0.05)

@@

We have gathered data on the relative speed and accuracies of \textbf{{\color{black}243}} bioinformatic methods from \textbf{{\color{black}43}} benchmarks that were published between 2005 and 2016. We show that there is an under-representation of software that has both intermediate levels of accuracy and speed. There may be a number of factors that drive this phenomena. One possible explanation for this is that benchmarkers select software that fills the extremes of each accuracy/speed niche. This seems unlikely since based upon our word frequency analysis the terms ``broad'', ``comprehensive'', ``inclusive'' and ``thorough'' are over-represented by between 200\% and 4,400\%.  A more likely explanation, in our opinion, is that bioinformatic software tools suffer from a form of publication-bias \cite{Boulesteix2015-am,Nissen:2016}. That is, the probability of a study being published is influenced by the results it contains \cite{sterling1959publication}. The community of developers, reviewers and editors appears to be unwilling to publish software that is neither the fastest nor the most accurate (we have anecdotal evidence to support this, N=1). If correct, this is unfortunate. Some problems that lack fast and accurate solutions force researchers to make unnecessary compromises in terms of accuracy and time. If our hypothesis that this underrepresentation is due to publication-bias is valid then why is there an enrichment of slow and inaccurate software (P-value=0.05, empirical distributions from permutation tests)? How could these methods be published? We have found that slow and inaccurate software is generally published earlier than alternative methods (P=0.007, one-tailed Wilcoxon test), therefore comparisons were not required to publish these tools and the accuracies may have been over-optimistically reported \cite{Boulesteix2015-nc}.

% 0.91	0.0004054328	0.0002104564	broad             100*(0.0004054328/0.0002104564) = 192.6446
% 2.08	0.0028380296	0.0006656728	comprehensive     100*(0.0028380296/0.0006656728) = 426.34
% 2.23	0.0002027164	0.0000352306	thorough          100*(0.0002027164/0.0000352306) = 575.3987
% 3.86	0.0002027164	0.0000046356	inclusive         100*(0.0002027164/0.0000046356) = 4373.035

We found that no commonly used proxy for study quality is correlated with software accuracy. Neither, author reputation, number of citations, journal impact, relative age or speed appear to be associated with accurate software. Linear mixed models of these values also fail to identify predictors of software accuracy, as do methods for combining P-values. A great deal of criticism has been leveled specifically at journal impact factors \cite{Editors2006-ue,Rossner2007-yk,Lariviere2010-kx,Brembs:2013}. Our finding show that this measure is also not reflective of software quality. But neither is the H5 index which is thought to be a more robust measures of journal impact \cite{Harzing2008-wb}. The poor relationship between accuracy and both author reputation and the number of citations is particularly troubling. Both are related to “word of mouth” and “previously used in a similar analysis” which a recent survey of researchers suggested is a major influence on the selection of software tools \cite{Loman2015-bw}. This implies that the recorded high citation rates for bioinformatic software  \cite{Perez-Iratxeta2007-lv,Van_Noorden2014-kc,Wren2016-xy} is more a reflection of user-friendliness and the Matthew Effect \cite{Lariviere2010-kx,Merton1968-cb}. The Matthew Effect is a biblical reference that can be paraphrased as “the rich get richer and the poor get poorer”. In this context, highly cited software is more likely to be used and cited again, irrespective of relative performance.

The lack of any relationship between software speed and accuracy is
surprising.  The slower software tools are overrepresented at both
high and low levels of accuracy
(Figure~\ref{fig:speedaccuracy}). Likewise accurate software was found
in both high and low speed ranges. A gedankenexperiment may be
sufficient to show that slow software is less thoroughly tested than
fast software. This is because typical academic software development
is an iterative process, where methods are refined by successive
rounds of coding, testing and evaluation \cite{Wilson2006-ih}. We can
assume that similar time-spans are spent on most projects, for
example, the span of a PhD degree or Postdoctoral Fellowship.  As a
consequence, slow software typically undergoes fewer development
cycles than fast methods, and is tested less, this may
result in less accurate and slower software. The fact that fast and
inaccurate software is relatively rare supports this argument.
%The
%lack of any further apparent relationship between speed and accuracy
%implies there are stronger influences on accuracy rankings than speed
%alone.

\section*{Conclusions}
Scientific progress is a slow process made by testing hypotheses using
appropriate positive and negative controls, in combination with
independent experiments and replication
\cite{Ioannidis2005-xh,Moonesinghe2007-kk}. We propose that software
and analysis tools should be no exception to these principles. The
continual increases in the complexity of software tools creates an
increased likelihood that software bugs are introduced
\cite{Darriba2015-sz}. Scientific software should therefore be
thoroughly tested by developers, however, the results of developer and
author-derived tests should be treated with caution due to a range of
potential conscious and unconscious biases \cite{Norel2011-cq}. Our
study shows that no commonly-used impact-based metrics are related to
software accuracy, therefore the only reasonable way to select
software tools is through software benchmarks.

We have found that slow and inaccurate software is typically published
early in the development of a field (see Figure S8). Subsequent
software tools may be high accuracy or high speed. However, as further
software tools become available, those that do not rank highly in
terms of accuracy or speed are under-represented in the
literature. Our conjecture is that this is due to a publication-bias
in the computational biology software literature, with publishers, editors
and reviewers implicitly requiring software to be notable in at least
one of these regards in order to be published. This hole in the
literature leaves an unfortunate gap upon which further software
refinements cannot be made.

The enrichment of slow and inaccurate software may be the primary
reason for masking the association between software accuracy and speed
that we had anticipated before embarking upon this study.

The software that could be used in building research workflows can
make use of software tools that are of intermediate accuracy and
speed. These may prove to be useful rapid pre-filters for more
accurate, yet computationally-demanding approaches. These could also
be used when datasets are too large to be processed by more accurate
software tools, i.e. when accurate tools are CPU or memory intensive
\cite{Lowe1997-hq,Weinberg2006-tm}. These tools may also serve as an
indication that some approaches may not be worth pursuing further.

We propose that the full spectrum of software tool accuracies and
speeds serve a useful purpose to the research community. Like negative
results, if honestly reported, illustrate to the research community
that certain approaches are not practical research avenues
\cite{Ioannidis2005-xh,Workman1999-au,Rivas2000-fb}. The classes of
tools that we find in the under-represented accuracy and speed region
include some of the most widely used and useful tools in
bioinformatics. These include the homology search tools NCBI-BLAST
\cite{Altschul1990-ht,Altschul1997-ga} and HMMER2 \cite{Eddy:2009} and
genome assembly and mapping tools such as AbySS \cite{Simpson:2009},
MAQ \cite{Li:2008} and Newbler \cite{Margulies:2005}. The current
practises of publishers, editors, reviewers and authors of software
tools therefore deprive our community of tools for building effective
and productive workflows.

@@

We have shown that accurate software is not necessarily the most
recently released method or the product of high profile lab groups or
selected by high impact journals. Software that is widely used or is
either slow or fast is also not necessarily the most
accurate. Therefore, accurate software may be the product of features
we have not been able to capture such as scientific inspiration, hard work,
sound method testing, technical ability and
education levels in software development. % \cite{Wilson2006-ih}.

Finally, we think that the field of computational biology could
benefit from embracing an increased number of independent software
comparison studies \cite{Boulesteix2013-vb}, in parallel with the
popular challenge-based benchmarks such as CASP, Assemblathon,
Alignathon, DREAM and RGASP \cite{Boutros2014-zm}. We are hopeful that
this, along with steps to reduce the publication-bias we have
described, will reduce the over-optimistic and misleading reporting of
method accuracy
\cite{Boulesteix2010-te,Jelizarow2010-zf,Norel2011-cq}.

@@

\section*{Acknowledgements}

The authors acknowledge the valued contribution of invaluable
discussions with Anne-Laure Boulesteix, Shinichi Nakagawa, Suetonia
Palmer and Jason Tylianakis. Michael A. Black, Murray Cox, Raquel
Norel, Alexandros Stamatakis, Jens Stoye,Tandy Warnow, provided
valuable feedback on drafts of the manuscript.

PPG is supported by a Rutherford Discovery Fellowship,
administered by the Royal Society of New Zealand.
